Efficient ancestry and mutation simulation with msprime 1.0

### 2021-08-09 paper.tex
Update paper.tex
added my affiliation

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{listings}
\usepackage[english]{babel}%
\usepackage[T1]{fontenc}%
\usepackage[utf8]{inputenc}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{dsfont}
\usepackage{verbatim}%
\usepackage{environ}%
\usepackage[right]{lineno}%
\usepackage{nameref}
\usepackage{tikz}
%\usepackage{showkeys}

@@

\title{Msprime 1.0: an efficient and extensible coalescent simulation framework}
\author{Author list to be filled in
}
% \address{

%%Affiliations:
% Franz Baumdicker:
% Cluster of Excellence "Controlling Microbes to Fight Infections", Mathematical and Computational Population Genetics, University of Tübingen, Germany

% Consuelo D. Quinto-Cortés
%National Laboratory of Genomics for Biodiversity (LANGEBIO), Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico

% Peter L. Ralph
% Institute of Ecology and Evolution, Departments of Mathematics and Biology, University of Oregon, USA

% \author[n]{Graham Gower}
% \affil[n]{Lundbeck GeoGenetics Centre, Globe Institute, Faculty of Health and Medical Sciences, University of Copenhagen, Denmark}

@@

\subsection*{Recombination}
\label{sec-recombination}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/gc-perf}
\end{center}
\caption{\label{fig-gc-perf}Comparison of simulation performance
using \msprime, \SimBac, and \FastSimBac\ for varying sample sizes,
with parameters roughly equivalent to the current estimate
for \textit{E.~coli}~\citep{lapierre2016the}:
a 4.5Mb genome with scaled gene conversion rate of 0.015 and
a mean tract length of 500.
We report (A) the total CPU time and (B) maximum memory usage
averaged over 5 replicates (Intel Xeon E5-2680 CPU).
We did not run \SimBac\ beyond first two data points because
of the very long running times.}
% The genome length was set to 4.5 Mb, the scaled per site
% gene conversion rate was set to $N_e g = \gamma = 0.015$ and
% the mean gene conversion tract length was set to $L = 500$, which is close to
% estimated parameters for \textit{E.~coli}, i.e.\ an effective populations size
% $N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
% with mean tract length $L = 542$~\citep{lapierre2016the}.}
\end{figure}

@@

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
as a representation of the coalescent with recombination
stochastic process as a graph.
This formulation is complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of recombination. In Griffiths' ARG formulation,
a realisation of the coalescent with recombination is a graph in which
vertices represent common ancestor or recombination events, and edges represent
lineages. There is the ``big'' ARG, in which we track lineages arising out of
recombinations regardless of whether they carry ancestral
material~\citep{ethier1990two},
% and
% consequently wait an exponentially large time until all sampled lineages
% coalesce into a single ancestor~\citep{ethier1990two},
and the ``little'' ARG in which we only track genetic ancestors.
Over time, the interpretation of the term
has shifted away from the original definition as a stochastic
process, to a representation of a \emph{particular} genetic ancestry
using common ancestor and recombination
events~\citep[e.g.][]{minichiello2006mapping}.
More recently, ARGs have essentially come to mean any graph-like representation
of genetic ancestry; for example,
\cite{mathieson2020ancestry} informally define an ARG as the subset of an
individual's pedigree from which they have inherited ancestral material.
% Do we include this sentence? It's true, I'm sure: someone new to the field reads
% Iain and Aylwyn's paper, and goes, hmm, this ARG thing sounds interesting,
% let's follow up on that Griffiths reference... Maybe this is part of the
% reason why ARGs have such a fearsome reputation for being difficult:
% there's a basic confusion about what they actually *are*.
Such usage is entirely reasonable, but referring readers back to the
original Griffiths papers does little to clarify.

@@

    \end{tikzpicture}
% \begin{center}
% $\equiv$
% \end{center}
% \end{column}
% \begin{column}{5cm}
\includegraphics[width=3.8cm]{illustrations/arg-ts.pdf}
\includegraphics[width=3.8cm]{illustrations/arg-ts-simplified.pdf}
% \includegraphics[width=3.8cm]{figures/arg.pdf}

\end{center}
\caption{\label{fig-arg} (A) A simple ARG in which a recombination
occurs at position 0.3; (B) the equivalent topology depicted as a tree
sequence, including the recombination node; (C) the same tree sequence
topology ``simplified'' down to its minimal tree sequence representation.
% (D) the fraction of nodes in a tree sequence that are concerned with
% extra ARG information in a tree sequence returned using \msprime's
% \texttt{record\_full\_arg} mode}
}
\end{figure}

For our purposes, an ARG is a realisation of the coalescent with
recombination, in the Griffiths (little ARG) sense.
As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
works by dynamically generating and traversing a little ARG.
The graph is not explicitly represented in memory, but is partially
present through the extant lineages and the ancestral material they carry
over time. We do not output the graph directly, but
rather store the information required to recover the genealogical
history as nodes and edges in a tree sequence (see the \nameref{sec-ts} section),
or ``coalescence records'' in original terminology~\citep{kelleher2016efficient}.
This is far more efficient than outputting the simulated ARG in its entirety.
For a given scaled recombination rate $\rho$
(setting aside the dependency on the sample size $n$)
we know from Eq.~\eqref{eqn-hsw-bound} that the number nodes
in an ARG is $O(\rho^2)$,
whereas the size of the tree sequence encoding is
$O(\rho)$~\citep{kelleher2016efficient}.
This difference between a
quadratic and a linear dependency on $\rho$ is profound, and shows why
large simulations cannot output an ARG in practice.

Although by default \msprime\ outputs tree sequences that contain full information
about the genealogical trees, their correlation structure along the genome,
and the ancestral haplotypes on which coalescences occurred, some information
is lost in this mapping down from ARG space to the minimal tree sequence
form. In particular, we lose
information about ancestral haplotypes that were common ancestors but
in which no coalescences occurred, and also information about the precise time
and genomic location of recombination events. In most cases, such
information is of little relevance as it is in principle unknowable,
but there are occasions such as visualisation or computing likelihoods (see
below) in which is useful. We therefore provide
the \texttt{record\_full\_arg} option in \msprime\
to store a representation of the complete ARG traversed during simulation.
This is done by storing extra nodes (marked with specific flags, so they
can be easily identified later) and edges in the tree sequence
(Fig.~\ref{fig-arg}).
One situation in which a record of the full ARG is necessary is when we
wish to compute likelihoods during inference.
The likelihood is a central quantity in evaluating the plausibility of a putative
ancestry as an explanation of DNA sequence data, both directly through
e.g.~approaches based on maximum likelihood, and as an ingredient of
methods such as
Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
We provide functions to compute the likelihood of ARG realisations
and mutational patterns under
the standard coalescent and infinite sites mutation model.
See the Appendix for details on these likelihood calculations.

>>>>

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{listings}
\usepackage[english]{babel}%
\usepackage[T1]{fontenc}%
\usepackage[utf8]{inputenc}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{dsfont}
\usepackage{verbatim}%
\usepackage{environ}%
\usepackage[right]{lineno}%
\usepackage{nameref}
\usepackage{tikz}
%\usepackage{showkeys}

@@

\title{Efficient ancestry and mutation simulation with msprime 1.0}
% First authors
%% FIXME affiliations are all broken
\author[1,$\star$]{Franz Baumdicker}
\author[2,$\star$]{Gertjan Bisschop}
\author[3,$\star$]{Daniel Goldstein}
\author[4,$\star$]{Graham Gower}
\author[5,$\star$]{Aaron P. Ragsdale}
\author[6,$\star$]{Georgia Tsambos}
\author[7,$\star$]{Sha Zhu}

% Middle Authors
\author[8]{Bjarki Eldon}
\author[9]{Castedo E. Ellerman}
\author[10]{Jared G. Galloway}
\author[11]{Ariella L. Gladstein}
\author[12]{Gregor Gorjanc}
\author[13]{Guo, Bing}
\author[7]{Ben Jeffery}
\author[14]{Warren W. Kretzschmar}
\author[15]{Ivan Krukov}
\author[2]{Konrad Lohse}
\author[16]{Michael Matschiner}
\author[15]{Dominic Nelson}
\author[17]{Nathaniel Pope}
\author[18]{Consuelo D. Quinto-Cort\'es}
\author[10]{Murillo F. Rodrigues}
\author[19]{Kumar Suanack}
\author[20]{Thibaut Sellinger}
\author[21]{Kevin Thornton}
\author[22]{Hugo van Kemenade}
\author[7]{Anthony W. Wohns}
\author[7]{H. Yan Wong}

% Senior Authors
\author[15,$\dagger$]{Simon Gravel}
\author[10,$\dagger$]{Andrew Kern}
\author[23,$\dagger$]{Jere Koskela}
\author[10,24,$\dagger$]{Peter L. Ralph}

% Corresponding
\author[7,$\ddagger$]{Jerome Kelleher}

% Franz
\affil[1]{Mathematical and Computational Population Genetics, University of T\"ubingen}
% Gertjan, Konrad
\affil[2]{Institute of Evolutionary Biology, The University of Edinburgh}
% Daniel
\affil[3]{FIXME: DANIEL'S AFFILIATION}
% Graham
\affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
% Simon, Dom, Ivan
\affil[5]{FIXME AARON's AFFILIATION}
% Georgia
\affil[6]{Melbourne Integrative Genomics, School of Mathematics and Statistics,
    University of Melbourne}
% Jerome, Yan, Wilder, Joe
\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
    University of Oxford}

% Bjarki
\affil[8]{FIXME: BJARKI's AFFILIATION}
% Castedo
\affil[9]{FIXME: CASTEDO's AFFILIATION}
% Jared, Murillo, Peter, Andy
\affil[10]{Department of Biology and Institute of Ecology and Evolution,
    University of Oregon}
% Ariella
\affil[11]{Department of Genetics, University of North Carolina at Chapel Hill}
% Gregor
\affil[12]{The Roslin Institute and Royal (Dick) School of Veterinary Studies,
    University of Edinburgh}
% Bing
\affil[13]{FIXME: BING's AFFILIATION}
% Winni
\affil[14]{FIXME: WINNI's AFFILIATION}
% Simon, Dom, Ivan
\affil[15]{Department of Human Genetics, McGill University}
% Michael
\affil[16]{FIXME: MICHAEL's AFFILIATION}
% Nathaniel
\affil[17]{FIXME: NATHANIEL's AFFILIATION}
% Consuelo
\affil[18]{National Laboratory of Genomics for Biodiversity (LANGEBIO),
    Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
% Saunack
\affil[19]{FIXME: SAUNACK's AFFILIATION}
% Thibaut
\affil[20]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
% Kevin
\affil[21]{Ecology and Evolutionary Biology, University of California at Irvine}
% Hugo
\affil[22]{FIXME HUGO's AFFILIATION}
% Jere
\affil[23]{FIXME: JERE'S AFFILIATION}
% Peter (second affil)
\affil[24]{Department of Mathematics, University of Oregon}

\affil[$\star$]{Denotes shared first authorship, listed alphabetically}
\affil[$\dagger$]{Denotes shared senior authorship, listed alphabetically}
\affil[$\ddagger$]{Denotes corresponding author}

@@

\subsection*{Recombination}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/gc-perf}
\end{center}
\caption{\label{fig-gc-perf}Comparison of simulation performance
using \msprime, \SimBac, and \FastSimBac\ for varying sample sizes,
with parameters roughly equivalent to the current estimate
for \textit{E.~coli}~\citep{lapierre2016the}:
a 4.5Mb genome with scaled gene conversion rate of 0.015 and
a mean tract length of 500.
We report (A) the total CPU time and (B) maximum memory usage
averaged over 5 replicates (Intel Xeon E5-2680 CPU).
We did not simulate \SimBac\ after the first two sample sizes because
of the very long running times.}
% The genome length was set to 4.5 Mb, the scaled per site
% gene conversion rate was set to $N_e g = \gamma = 0.015$ and
% the mean gene conversion tract length was set to $L = 500$, which is close to
% estimated parameters for \textit{E.~coli}, i.e.\ an effective populations size
% $N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
% with mean tract length $L = 542$~\citep{lapierre2016the}.}
\end{figure}

@@

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
as a representation of the coalescent with recombination in terms of a graph.
This formulation was complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of recombination. In Griffiths' ARG formulation,
a realisation of the coalescent with recombination is a graph in which
vertices represent common ancestor or recombination events, and edges represent
lineages through which ancestral material flows.
We may consider the big ARG, in which we track lineages arising out of
recombinations regardless of whether they carry ancestral material and
consequently wait an exponentially large time until all sampled lineages
coalesce into a single ancestor~\citep{ethier1990two}, or the little ARG in
which we only track genetic ancestors, and wait until all marginal trees
have fully coalesced (like Hudson's algorithm).

@@

    \end{tikzpicture}
% \begin{center}
% $\equiv$
% \end{center}
% \end{column}
% \begin{column}{5cm}
\includegraphics[width=3.8cm]{illustrations/arg-ts.pdf}
\includegraphics[width=3.8cm]{illustrations/arg-ts-simplified.pdf}
\includegraphics[width=3.8cm]{figures/arg.pdf}

\end{center}
\caption{\label{fig-arg} (A) A simple ARG in which a recombination
occurs at position 0.3; (B) the equivalent topology depicted as a tree
sequence, including the recombination node; (C) the same tree sequence
topology ``simplified'' down to its minimal tree sequence representation;
(D) the fraction of nodes in a tree sequence that are concerned with
extra ARG information in a tree sequence returned using \msprime's
\texttt{record\_full\_arg} mode}
\end{figure}

However, this original definition of the ARG in terms of a stochastic process
is rarely what is meant when the term is currently used. The term has come to
essentially mean any graph-like representation of genetic ancestry in the presence
of recombination; for example,
\cite{mathieson2020ancestry} informally define an ARG as the subset of an
individual's pedigree from which they have inherited ancestral material.
Others state that an ARG is equivalent to the sequence of marginal
% Also SARGE paper
trees along the genome~\citep{yang2016rapidly,mirzaei2016rent}

Such
imprecision is unfortunate, since there are many ways to encode genetic
ancestry as a graph, and referring to the original Griffiths papers does little to
clarify. As we see below, there are situations in which the precise information
stored in the graph is important, and conflating a high-level
discussion of genetic ancestry with this mathematical object is not
helpful. The point of technical terms is to provide clarity and
precision on what is being discussed, and it is now essentially
impossible to know what is meant by the term ``ARG'';
we therefore use the term ``genetic ancestry'' in high-level
discussions.

For our purposes, an ARG is a realisation of the coalescent with
recombination, in the Griffiths (little ARG) sense.
As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
works by dynamically generating and traversing a little ARG.
The graph is not explicitly represented in memory, but is partially
present through the extant lineages and the ancestral material they carry
over time. Since the ARG is a large and unwieldy object
(Fig.~\ref{fig-arg}) we do not use it as output, but
rather store the information required to recover the genealogical
history as nodes and edges in a tree sequence (see the \nameref{sec-ts} section),
or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.

Although by default \msprime\ outputs tree sequences that contain full information
about the genealogical trees, their correlation structure along the genome,
and the ancestral haplotypes on which coalescences occurred, some information
is lost in this mapping down from ARG space to the minimal tree sequence
form. In particular, we lose
information about ancestral haplotypes that were common ancestors but
in which no coalescences occurred, and also information about the precise time
and genomic location of recombination events. In most cases, such
information is of little relevance as it is in principle unknowable,
but there are occasions such as visualisation or computing likelihoods (see
below) in which is useful. We therefore provide
the \texttt{record\_full\_arg} option in \msprime\
to store a representation of the complete ARG traversed during simulation.
This is done by storing extra nodes (marked with specific flags, so they
can be easily identified later) and edges in the tree sequence.
Fig.~\ref{fig-arg} shows that this extra ARG information
quickly comes to dominate the output as we increase the sequence length,
and the resulting tree sequence files are many times larger
than the standard minimal tree sequence representation.

One situation in which a record of the full ARG is necessary is when we
wish to compute likelihoods during inference.
The likelihood is a central quantity in evaluating the plausibility of a putative
ancestry as an explanation of DNA sequence data, both directly through
e.g.~approaches based on maximum likelihood, and as an ingredient of
methods such as
Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
We provide functions to compute the likelihood of ARG realisations
(encoded using the representation of the \texttt{record\_full\_arg} option)
and mutational patterns under
the standard coalescent and infinite sites mutation model.
See the Appendix for details.
